---
title: "Supplemental Figure 4 - Differential Abundance Analysis"
output: 
  pdf_document: default
  github_document: 
    toc: true
urlcolor: blue
# Created by Timothy Keyes
# Reformatted by Alea Delmastro
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, dpi = 300, warning = FALSE, message = FALSE)
```

# 0. Libraries

```{r}
# libraries
library(diffcyt)
library(tidyverse)
library(tidymodels)
library(lme4)

# source utils
source('/Users/aleadelmastro/Downloads/pap_utils.R', echo = FALSE)
```

# 1. Setup

```{r}
# paths 
path <- "/Users/aleadelmastro/Library/CloudStorage/GoogleDrive-alead@stanford.edu/.shortcut-targets-by-id/1GXGQlL1W-w8iJENEwcrx7NBGohHlnlCY/PAP/PAP_MIBI/Cohort/MIBI_data/all_points"

data_path <- paste(path, "/celltable_05032024.csv", sep = "")

metadata_path <- paste(path, "/fov_labels.csv", sep = "")
```

`r pagebreak()`

# 2. Read in data

```{r}
# metadata 
metadata <- 
  metadata_path %>% 
  read_csv() %>% 
  rename(fov_id = point) 

# mibi data
mibi_data <- 
  data_path %>% 
  read_csv() %>% 
  rename(
    fov_id = point, 
    cell_id = label, 
    cluster_id = pixelfreq_hclust_cap,
    cluster_name = name, 
    centroid_x = `centroid.0`, 
    centroid_y = `centroid.1`
  )
```

`r pagebreak()`

# 3. Pre-process data

```{r}
# join the mibi single-cell data with the FOV metadata
mibi_data <- 
  mibi_data %>% 
  left_join(metadata)

# get fov ids for uninvolved regions
uninvolved_fovs_in_pap_patients <- metadata[metadata$FOV_Category == "uninvolved",]$fov_id
```

`r pagebreak()`

# 4. Basic data summary

## Number of unique FOVs in each condition

```{r}
mibi_data %>% 
  distinct(FOV_Category, fov_id) %>% 
  count(FOV_Category, name = "num_fovs") %>% 
  arrange(-num_fovs)
```

## Number of unique patients in each condition

```{r}
#Get Patient Number:
mibi_data$Patient_num <- sub("\\..*", "", mibi_data$Patient_ID)

mibi_data %>% 
  distinct(Tissue_Category, Patient_num) %>% 
  count(Tissue_Category, name = "num_patients") %>% 
  arrange(-num_patients)
```

## Number of unique cells in each condition

```{r}
mibi_data %>% 
  count(FOV_Category, name = "num_cells") %>% 
  arrange(-num_cells)
```

## Number of cells for each patient 

```{r}
mibi_data %>% 
  count(Patient_num, name = "num_cells") %>% 
  arrange(-num_cells)
```


## Number of cells in each FOV

```{r}
mibi_data %>% 
  count(fov_id, name = "num_cells") %>% 
  arrange(-num_cells)
```

`r pagebreak()`

# 6. Differential Abundance Analysis - within patients

In addition to the between-patients comparisons, we can also run another set of comparisons that leverages a within-subjects design to increase statistical power. As it turns out, for several of our sJIA-daPAP samples, one of the FOVs collected was annotated as a "uninvolved" section of tissue relative to the others (which had more of the hallmark histopathological features of sJIA-daPAP). We can compare the abundance of each of our immune cell subpopulations within the same patients by comparing the "uninvolved" FOV to the other FOVs taken from the same patient. In this case the GLMM we're using is the following: 


$$
logit(p_{ij}) = log(\frac{p_{ij}}{1 - p_{ij}}) = \beta_0 + \alpha_i + \beta_1 X_{j_{diseased}}, 
$$
where $X_{j_{diseased}}$ is an indicator variable representing if FOV j comes from an "involved" area of the sJIA-daPAP tissue or not (1 if yes; 0 otherwise). Thus, $\beta_1$ represents the difference in log-odds for a given cell subtype between the involved and uninvolved FOVs within a given patients (controlling for patient-to-patient variability, which is represented by $\alpha_i$).  

## Processing

```{r}
# find patients who had at least one "healthy" FOV
interesting_patients <- 
  mibi_data %>% 
  filter(fov_id %in% uninvolved_fovs_in_pap_patients) %>% 
  distinct(Patient_num) %>% 
  pull(Patient_num)

interesting_patients
```

```{r}
# filter only the patients with "healthy-looking" FOVs from the full dataset
paired_patients <- 
  mibi_data %>% 
  filter(Patient_num %in% interesting_patients) %>% 
  # annotate FOVs that are "healthy-looking" according to our pathologist
  mutate(
    fov_condition = 
           if_else(fov_id %in% uninvolved_fovs_in_pap_patients, "uninvolved", "involved")
  )

```

## Statistical analysis

```{r, message = FALSE, warning = FALSE}
paired_daa_results <- 
  paired_patients %>% 
  filter(Tissue_Category == "sJIA-daPAP") %>% 
  pap_perform_daa(
    data_tibble = ., 
    sample_col = fov_id, 
    cluster_col = cluster_name, 
    fixed_effect_cols = fov_condition, 
    random_effect_cols = Patient_num, 
    include_observation_level_random_effects = FALSE
  )

paired_daa_results %>% 
  pluck("da_results") %>% 
  topTable(all = TRUE) %>% 
  as_tibble() %>% 
  mutate(significant = if_else(p_adj < 0.05, "*", "")) %>% 
  arrange(p_adj) %>%  
  knitr::kable()
```


From these results, we can see that there are several immune cell subtypes that, when using a paired design, we find are enriched in parts of the sJIA-daPAP lung that actually show histopathological signs of disease compared to paired parts of the sJIA-daPAP lung that do not show histopathological signs of disease.

We can visualize these differences below. 

## For visualization

```{r}
paired_p_values <- 
  paired_daa_results %>% 
  pluck("da_results") %>% 
  topTable(all = TRUE) %>% 
  as_tibble() %>% 
  mutate(
    significant = if_else(p_adj < 0.05, "*", ""), 
    new_cluster_name = if_else(significant == "*", str_c(cluster_id, "*"), as.character(cluster_id))
  ) %>% 
  arrange(p_adj) %>% 
  rename(cluster_name = cluster_id)

sig_clusters <- 
  paired_p_values %>% 
  filter(significant == "*") %>% 
  pull(cluster_name)

# calculate the number of fovs used for each patient in each "condition"
num_fov_tibble <- 
  paired_patients %>% 
  distinct(fov_id, Patient_ID, fov_condition) %>% 
  count(Patient_ID, fov_condition, name = "num_fovs")

paired_plot_data <- 
  paired_patients %>% 
  mutate(
    cluster_name = 
      factor(cluster_name, levels = pull(paired_p_values, cluster_name)) , 
  ) %>% 
  count(cluster_name, Patient_ID, fov_id, fov_condition, .drop = FALSE) %>%
  group_by(fov_id) %>% 
  mutate(
    total_fov_cells = sum(n),
    prop = n / total_fov_cells
  ) 
```

## Plot Differential Abundance

```{r}
paired_plot_data %>% 
  left_join(paired_p_values) %>% 
  mutate(new_cluster_name = fct_reorder(new_cluster_name, p_adj)) %>%  
  drop_na(new_cluster_name) %>% 
  select(Patient_ID, fov_condition, new_cluster_name, prop, p_adj) %>% 
  pivot_wider(
    names_from = fov_condition, 
    values_from = prop
  ) %>% 
  group_by(new_cluster_name) %>% 
  summarize(across(c(uninvolved, involved), mean, na.rm = TRUE)) %>% 
  mutate(
    fc = involved / uninvolved, 
    log2_fc = log(fc, base = 2)
  ) %>% 
  left_join(paired_p_values) %>% 
  mutate(
    neg_log_p_val = -log(p_adj),
    cluster_type = 
      case_when(
        p_adj > 0.05 ~ "Not significant", 
        fc > 1       ~ "Increased", 
        fc < 1       ~ "Decreased"
      )
  ) %>% 
  ggplot(aes(x = log2_fc, y = neg_log_p_val, fill = cluster_type)) + 
  geom_hline(yintercept = -log(0.05), color = "black", linetype = "dashed") + 
  geom_vline(xintercept = 0, color = "black", linetype = "dashed") +
  geom_point(shape = 21, size = 2.5) + 
  ggrepel::geom_label_repel(
    aes(label = cluster_name), 
    size = 2, 
    color = "black", 
    show.legend = FALSE
  ) + 
  scale_y_continuous(limits = c(NA, 20), oob = scales::oob_squish_any) + 
  labs(
    subtitle = "Differentially abundant clusters in involved vs. uninvolved regions of sJIA-daPAP lung", 
    x = "log2FC", 
    y = "-log(p-value)",
    fill = NULL, 
    caption = "Increased/decreased refer to abundance of clusters in\ndiseased regions relative to healthy regions of lung tissue"
  ) + theme_minimal() + theme(panel.border = element_rect(colour = "black", fill = NA, size = 1), axis.ticks.length = unit(0.1, "cm"),
    axis.ticks = element_line(color = "black", size = 0.5))

ggsave(paste(path, "/Sub-panels/SupplementalFigures/SuppFig4_DA_plot.pdf", sep = ""), width = 6, height = 4, units = "in")
```










